IRS County-to-County Migration Data, 1990-2010

### 2018-08-24 manuscript_file.Rmd
ready for submission

---
output: 
  pdf_document:
    number_sections: true
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "IRS County-to-County Migration Data, 1990-2010"
thanks: "The data and code that supports the creation of this data are available in the Supplementary Materials and online at  https://github.com/blinded/for-peer-review."
author:
  # - name: Mathew E. Hauer ^1,2^*
  #   affiliation: Florida State University
  # - name: James M. Byars ^3^
  #   affiliation: University of Georgia
# abstract: ""
keywords: "Migration, IRS, Data, R"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: single
bibliography:  LATEX/mybibfile.bib
biblio-style: apsr
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage{multirow}
- \usepackage{tabularx}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[CE,CO]{IRS Migration Data}

@@

**BACKGROUND**: The Internal Revenue Service's (IRS) county-to-county migration data are an incredible resource for understanding migration in the United States. Produced annually since 1990 in conjunction with the US Census Bureau, the IRS migration data represent 95 to 98 percent of the tax filing universe and their dependents, making the IRS migration data one of the largest sources of migration data. However, any analysis using the IRS migration data must process at least seven legacy formats of these public data across more than 2000 data files -- a serious burden for migration scholars.

**OBJECTIVE**: To produce a single, flat data file containing complete county-to-county IRS migration flow data and to make the computer code used to process the migration data freely available.

@@

In the United States, subnational migration flow is available from three primary sources, depending on time period: the Decennial Census, the American Community Survey, and the Internal Revenue Service's (IRS) county-to-county migration data. The IRS migration data are a pioneering use of administrative records to estimate demographic processes and are available on an annual basis since 1990. Because of the annual availability, relatively long time series, large universe due to the administrative records, and long history of use, the IRS data are an attractive data source for conducting migration research in the United States (e.g. [@curtis2015recovery;@molloy2011internal;@shumway2001spatial;@frey2009great]). Unfortunately, these data exist in seven legacy formats, split between 2,000+ data files making analysis with this data rather burdensome and has likely hindered the widespread adoption of this valuable resource for US migration scholarship.

@@

We have attempted to introduce as little post-processing as possible to process the data into a common format. US Counties are fairly stable geographic units but some changes in county boundaries, names, and FIPS codes due occasionally occur^[The Federal Information Processing Standard Publication (FIPS) is a 5-digit code used to uniquely identify US counties and county equivalents.]. To try and keep as close to the original data fidelity as possible, we did not recode any geographic changes and present the IRS migration data as-is. For instance, Broomfield County, Colorado (FIPS 07014) was created out of parts of Adams, Boulder, Jefferson, and Weld counties in 2001 and thus has data only after 2002. Uses should be aware of any changes in county boundaries, county names, or FIPS changes that could substantially alter any analysis of this data^[More detailed information about county boundary, name, or FIPS changes can be found at the following locations https://www.census.gov/geo/reference/county-changes.html
http://www.nber.org/asg/ASG_release/County_City/FIPS/FIPS_Changes.pdf
https://www.cdc.gov/nchs/data/nvss/bridged_race/County_Geography_Changes.pdf
https://www.ddorn.net/data/FIPS_County_Code_Changes.pdf].

The following document is organized as follows. First, we describe the IRS county-to-county migration data to provide an overview of the data for scholars who might be unfamiliar with the IRS migration data. Second, we provide usage notes that provide important information that may assist other researchers who use our data. Third, we describe our single, flat, standardized file and document important nuances in the raw IRS migration data. Finally, we describe parts of the R code used to download the IRS migration data and process it into a common format.

The IRS migration data are an incredible tool for understanding migration. By providing these data in a readily available format and the subsequent open-source computer code used to process these data, we hope to facilitate their use in descriptive, exploratory, and analytical analyses of migration in the United States through the use of administrative data. This is data is particularly useful for understanding migration as a spatial entity and for investigating the evolution of migration systems over time.

@@

The annual series of county-to-county migration data cover 95 to 98 percent of the tax filing universe (or approximately 87% of US households [@molloy2011internal]) and their dependents making these data the largest migration data source for count flows between counties in the United States. But since the IRS derives migration information from tax filings, those who do not file taxes are most likely to be underrepresented in the migration data [@gross2005internal; @dewaard2016population], namely undocumented populations, the poor, the elderly, and college students [@gross2005internal]. However, the overwhelming majority of householders file US tax returns in the United States [@molloy2011internal].

@@

Between 1990 and 2010, the IRS processed the county-to-county migration data using the same procedures. However, in 2011 the IRS introduced a new method for processing the migration data and introduced "enhancements" to improve the overall quality of the data [@pierce2015soi]. The IRS introduced three major changes. First, they began basing migration on a full year of data as opposed to a partial year of data. To meet Census Bureau deadlines, the IRS processed all income tax returns filed before the end of September and did not process the returns that were filed between the end of September and the end of the calendar year. Beginning with migration year 2011, the IRS included the approximately 4% of returns that are filed between the end of September and December 31, allowing the IRS to produce a full calendar years' worth of migration. Second, the IRS improved the year-to-year matching, increasing the number of matched returns by 5 percent. Third, the IRS began tabulating gross migration at the US State level by size of adjusted gross income (AGI) and the age of the primary taxpayer.

These changes to the processing of returns create a break in the historic time series. For this reason, we limit the data we process to the period 1990-2010, the last year before the new processing rules. If a scholar wishes to processes any IRS migration data after 2010, the R code that we provide can be easily adapted to do so.

@@

Users should also be mindful of possible geographic changes to county boundaries that could impact the data.

The county migration data we present come from the `exemptions` field of the IRS migration data. The original IRS migration data contains two consistent fields across all years of data: a `returns` field and an `exemptions` field. Returns are the number of tax returns filed while exemptions are a proxy for the members of the household. This was done to better mimic the number of individuals migrating rather than the number of households.

@@

# Data Processing
The IRS migration data for the period 1990-2010 are available in seven legacy formats. **Table 2** summarizes some of the similarities and differences in these formats. For every year, the IRS publishes approximately 104 data files. (52 state entities by in/out-migration. Some years contain .csv and .dat summary files.) The underlying file organization, file format, naming schema, and coding can differ between these legacy formats. Migration years 1990 and 1991 are available as fixed-width text files, while 1992-2010 are available as excel files. For years 1990-2003, the IRS separated in/out migration into separate folders while 2004-2010 are published in a single folder. Each legacy format utilizes a different file naming scheme as well, making pattern matching of file names (called grepping) difficult. Importantly, the IRS treats non-migrants and total migrants differently in the seven legacy formats. For 1990 and 1991, the IRS simply has a field that reads "County Non-Migrants" for non-migrants; for 1992-1994, the IRS introduced a State code 63 but two difference County codes (010 for 1992 and 1994 and 050 for 1993) creating a 5-digits FIPS code of 63010 or 63050. After 1995, the IRS smartly set the origin FIPS equal to the destination FIPS for non-migrants. Lastly, Total Migrants are treated differently too. For 1990 and 1991, the destination field simply reads "Total Migration." For 1992-1994, the IRS introduced a State Code 00 and county code 001 for total migrants. After 1995, the IRS used State Code 96 and county Code 000 for a combined 5-digit FIPS code of 96000.

The differences described above and in **Table 2** are only some of the differences that are of interest to the data we produce here. Total Migrants, ie FIPS 96000 for migration data after 1995, is also broken down into Total Mig - US (FIPS 97000), Total Mig - US Same State (FIPS 97001), Total Mig - US Diff St (FIPS 97003), and Total Mig - Foreign (FIPS 98000). The IRS did not code these migration flows in this manner for all years, and in some cases (such as Total Mig - Foreign) migration flows are not reported. For simplicity and data continuity purposes, we simply create a new origin/destination (FIPS 99999) that contains all unspecified migration flows. We do this by subtracting the number of enumerated migrants (the migration flows with greater than 10 migrants) from the total number of migrants. This way, the sum of all enumerated migrants in our dataset will equal the total number of migrants in the IRS dataset.

@@

The R code used to produce these data is available in the **Supplementary Materials** and can also be found in an online repository^[https://github.com/blinded/for-peer-review]. The code makes use of multi-core processing to speed up computation time. There are three main sections in the code: A setup section; a data download section; and a data processing section. The final flat file, `county_migration_data.txt`, contains the # of exemptions and can be either downloaded at github or produced by running the R code.

@@

## Data Download
The script `001-download_data.R` will download and unzip the migration data from the IRS' websites into a folder standardized format into subdirectory `MigData/.` The IRS data is in two primary formats: 1990-2003 and 2004-onward. There are eight files that the IRS includes in their zip archives that contain no data (these are in years 1998, 1999, 2000, and 2001). After being downloaded and unzipped, these files are then deleted. If they are not deleted, they will cause the subsequent `for loops` to fail in the next section. These files do not contain any migration information, their names suggest they represent aggregation of migration flows (for example 'co990usi.xls' suggests county 99-2000 US in-migration), and we are unsure exactly why these files are included or their purpose.

>>>>

---
output: 
  pdf_document:
    number_sections: true
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "IRS County-to-County Migration Data, 1990-2010"
thanks: "The data and code that supports the creation of this data are available in the Supplementary Materials and online at  https://github.com/blinded/for-peer-review."
author:
  # - name: Mathew E. Hauer ^1,2^*
  #   affiliation: Florida State University
  # - name: James M. Byars ^3^
  #   affiliation: University of Georgia
# abstract: ""
keywords: "Migration, IRS, Data, R"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography:  LATEX/mybibfile.bib
biblio-style: apsr
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
# - \usepackage{fancyhdr}
- \usepackage{multirow}
- \usepackage{tabularx}
# - \pagestyle{fancy}
- \pagenumbering{gobble}
# - \fancyhead{}
# - \fancyhead[CE,CO]{IRS Migration Data}

@@

**BACKGROUND**: The Internal Revenue Service's (IRS) county-to-county migration data are an incredible resource for understanding migration in the United States. Produced annually since 1990 in conjunction with the US Census Bureau, the IRS migration data represent 95 to 98 percent of the tax-filing universe and their dependents, making the IRS migration data one of the largest sources of migration data. However, any analysis using the IRS migration data must process at least seven legacy formats of these public data across more than 2000 data files -- a serious burden for migration scholars.

**OBJECTIVE**: To produce a single, flat data file containing complete county-to-county IRS migration flow data and to make the computer code to process the migration data freely available.

@@

In the United States, subnational migration flow data is available from three primary sources depending on the time period: the Decennial Census, the American Community Survey, and the Internal Revenue Service's (IRS) county-to-county migration data. The IRS migration data are a pioneering use of administrative records to estimate demographic processes and are available on an annual basis since 1990. Because of the annual availability, relatively long time series, large universe due to the administrative records, and long history of use, the IRS data are an attractive data source for conducting migration research in the United States (e.g. [@curtis2015recovery;@molloy2011internal;@shumway2001spatial;@frey2009great]). Unfortunately, these data exist in seven legacy formats, split between 2,000+ data files making analysis with this data rather burdensome and has likely hindered the widespread adoption of this valuable resource for US migration scholarship.

@@

We have attempted to introduce as little post-processing as possible to process the data into a common format. US Counties are fairly stable geographic units but some changes in county boundaries, names, and FIPS codes due occasionally occur^[The Federal Information Processing Standard Publication (FIPS) is a 5-digit code used to uniquely identify US counties and county equivalents.]. To try and keep as close to the original data fidelity as possible, we did not recode any geographic changes and present the IRS migration data as-is. For instance, Broomfield County, Colorado (FIPS 07014) was created out of parts of Adams, Boulder, Jefferson, and Weld counties in 2001 and thus has data only after 2002. Usees should be aware of any changes in county boundaries, county names, or FIPS changes that could substantially alter any analysis of this data^[More detailed information about county boundary, name, or FIPS changes can be found at the following locations https://www.census.gov/geo/reference/county-changes.html
http://www.nber.org/asg/ASG_release/County_City/FIPS/FIPS_Changes.pdf
https://www.cdc.gov/nchs/data/nvss/bridged_race/County_Geography_Changes.pdf
https://www.ddorn.net/data/FIPS_County_Code_Changes.pdf].

We organize the following document as follows. First, we describe the IRS county-to-county migration data to provide an overview of the data for scholars who might be unfamiliar with the IRS migration data. Second, we provide usage notes that provide important information that may assist other researchers who use our data. Third, we describe our single, flat, standardized file and document important nuances in the raw IRS migration data. Finally, we describe parts of the R code used to download the IRS migration data and process it into a common format.

The IRS migration data are an incredible tool for understanding migration. By providing these data in a readily available format and the subsequent open-source computer code used to process these data, we hope to facilitate their use in descriptive, exploratory, and analytical analyses of migration in the United States using administrative data. This data is particularly useful for understanding migration as a spatial entity and for investigating the evolution of migration systems over time.

@@

The annual series of county-to-county migration data cover 95 to 98 percent of the tax-filing universe (or approximately 87% of US households [@molloy2011internal]) and their dependents making these data the largest migration data source for count flows between counties in the United States. The IRS derives migration information from tax-filings making those who do not file taxes most likely to be underrepresented in the migration data [@gross2005internal; @dewaard2016population], namely undocumented populations, the poor, the elderly, and college students [@gross2005internal]. However, the overwhelming majority of householders file US tax returns in the United States [@molloy2011internal].

@@

Between 1990 and 2010, the IRS processed the county-to-county migration data using the same procedures. However, in 2011 the IRS introduced a new method for processing the migration data and introduced "enhancements" to improve the overall quality of the data [@pierce2015soi]. The IRS introduced three major changes. First, they began basing migration on a full year of data as opposed to a partial year of data. To meet Census Bureau deadlines, the IRS processed all income tax returns filed before the end of September and did not process the returns filed between the end of September and the end of the calendar year. Beginning with migration year 2011, the IRS included the approximately 4% of returns that are filed between the end of September and December 31, allowing the IRS to produce a full calendar years' worth of migration. Second, the IRS improved the year-to-year matching, increasing the number of matched returns by 5 percent. Third, the IRS began tabulating gross migration at the US State level by size of adjusted gross income (AGI) and the age of the primary taxpayer.

These changes to the processing of returns create a break in the historic time series. For this reason, we limit the data we process to the period 1990-2010, the last year before the new processing rules. If a scholar wishes to process any IRS migration data after 2010, the R code that we provide can be easily adapted to do so.

@@

Users should also be mindful of possible geographic changes to county boundaries that could affect the data.

The county migration data we present come from the `exemptions` field of the IRS migration data. The original IRS migration data contains two consistent fields across all years of data: a `returns` field and an `exemptions` field. Returns are the number of tax returns filed while exemptions are a proxy for the members of the household. We use the number of exemptions to better mimic the number of individuals migrating rather than the number of households.

@@

# Data Processing
The IRS migration data for the period 1990-2010 are available in seven legacy formats. **Table 2** summarizes some of the similarities and differences in these formats. For every year, the IRS publishes approximately 104 data files. (52 state entities by in/out-migration. Some years contain .csv and .dat summary files.) The underlying file organization, file format, naming schema, and coding can differ between these legacy formats. Migration years 1990 and 1991 are available as fixed-width text files, while 1992-2010 are available as excel files. For years 1990-2003, the IRS separated in/out migration into separate folders while 2004-2010 are published in a single folder. Each legacy format utilizes a different file naming scheme as well, making pattern matching of file names (called grepping) difficult. Importantly, the IRS treats non-migrants and total migrants differently in the seven legacy formats. For 1990 and 1991, the IRS simply has a field that reads "County Non-Migrants" for non-migrants; for 1992-1994, the IRS introduced a State code 63 but two different County codes (010 for 1992 and 1994 and 050 for 1993) creating a 5-digit FIPS code of 63010 or 63050. After 1995, the IRS smartly set the origin FIPS equal to the destination FIPS for non-migrants. Lastly, Total Migrants are treated differently too. For 1990 and 1991, the destination field simply reads "Total Migration." For 1992-1994, the IRS introduced a State Code 00 and county code 001 for total migrants. After 1995, the IRS used State Code 96 and county Code 000 for a combined 5-digit FIPS code of 96000.

The differences described above and in **Table 2** are only some of the differences that are of interest to the data we produce here. Total Migrants, ie FIPS 96000 for migration data after 1995, is also broken down into Total Mig - US (FIPS 97000), Total Mig - US Same State (FIPS 97001), Total Mig - US Diff St (FIPS 97003), and Total Mig - Foreign (FIPS 98000). The IRS did not code these migration flows in this manner for all years, and in some cases (such as Total Mig - Foreign) migration flows are not reported. For simplicity and data continuity purposes, we simply create a new origin/destination (FIPS 99999) that contains all unspecified migration flows. We do this by subtracting the number of enumerated migrants (the migration flows with greater than 10 migrants) from the total number of migrants. This way, the sum of all enumerated migrants in our dataset will equal the total number of migrants in the IRS dataset. And the sum of all migrants and non-migrants for any origin in a given year should roughly approximate the county population estimate for that year.

@@

The R code used to produce these data is available in the **Supplementary Materials** and in an online repository^[https://github.com/blinded/for-peer-review]. The code makes use of multi-core processing to speed up computation time. There are three main sections in the code: A setup section; a data download section; and a data processing section. The final flat file, `county_migration_data.txt`, contains the # of exemptions and can be either downloaded at github or produced by running the R code.

@@

## Data Download
The script `001-download_data.R` will download and unzip the migration data from the IRS' websites into a folder standardized format into subdirectory `MigData/.` The IRS data is in two primary formats: 1990-2003 and 2004-onward. The IRS includes eight files in their zip archives that contain no data (these are in years 1998, 1999, 2000, and 2001). After being downloaded and unzipped, these files are then deleted. If they are not deleted, they will cause the subsequent `for loops` to fail in the next section. These files do not contain any migration information, their names suggest they represent aggregation of migration flows (for example 'co990usi.xls' suggests county (co) years 1999-2000 (990) for US (us) in-migration (i)), and we are unsure exactly why these files are included or their purpose.
